% This is LLNCS.DEM the demonstration file of
% the LaTeX macro package from Springer-Verlag
% for Lecture Notes in Computer Science,
\documentclass[runningheads,a4paper]{llncs}
\usepackage{makeidx}  % allows for indexgeneration
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{listings}
\usepackage{mathpartir}
\usepackage{graphicx}
\usepackage{comment}
\usepackage{pifont}
\usepackage{enumitem}
\usepackage{url}
\lstset
{
	basicstyle = \ttfamily\tiny,
	breaklines = true,
	frame = single
}

\begin{document}
\mainmatter

\pagestyle{headings} 

\title{Automated networking in Meta-Casanova}

\titlerunning{Automated networking in Meta-Casanova}

\author{Francesco Di Giacomo\inst{1} \and Mohamed Abbadi\inst{1} \and Agostino Cortesi\inst{1}
	\and Pieter Spronck\inst{2}
	\and Giuseppe Maggiore\inst{3}}

\institute{Universita' Ca' Foscari, Venezia,\\
	\email{francesco.digiacomo@unive.it, cortesi@unive.it, mohamed.abbadi@unive.it}
	\and
	Tiburg University,
	\email{p.spronck@uvt.nl}
	\and
	Hogeschool Rotterdam,
	\email{maggg@hr.nl}}

\authorrunning{Francesco Di Giacomo et al.}
	
\maketitle


\abstract{Multiplayer support is a desirable feature in commercial games since it allows to extend the longevity of a title. The same importance can be found in several examples of serious games, as they are often used in training where several individuals must cooperate to reach a common goal. Having smaller resources than big entertainment companies, serious games developers cannot afford to develop ad-hoc networking code and need to rely on game engines. On the other hand, game engines require deep knowledge and do not overcome all the problems of hand-made networking implementations. For this reason, it is desirable to express networking behaviours in a dedicated language, or Domain-specific language, that provides the necessary abstractions, but this approach suffers from having to implement a compiler for the language, which is hard to extend. In this paper we provide an alternative approach based on meta-compilation defined in Metacasanova, a meta-compiler where it is possible to specify the formal semantics of the language. We show that using Metacasanova notably reduces the effort of developing a Domain-specific language while keeping the same advantage in terms of quality of code.}
	
\section{Introduction}
\input{Sections/introduction}

\section{Challenges of multiplayer games}
\input{Sections/problem}

\section{Networking architecture and semantics}
\input{Sections/architecture}

\section{Implementation in Metacasanova}
\input{Sections/implementation}

\section{Evaluation}
\input{Sections/evaluation}

\section{Conclusion}
\input{Sections/conclusion}

\bibliography{bibliography}
\bibliographystyle{plain}

\end{document}