\chapter*{List of publications}

\begin{enumerate}
\item Abbadi, Mohamed; Di Giacomo, Francesco; Maggiore, Giuseppe; Plaat, Aske; Orsini, Renzo; Spronck, Pieter \textit{Resource Entity Action: A generalized design pattern for RTS games} , Proceedings of the 8th International Conference on Computers and Games (CG2013), Springer Verlag.
\item Abbadi, Mohamed ; Di Giacomo, Francesco; Cortesi, Agostino; Spronck, Pieter; Costantini, Giulia; Maggiore, Giuseppe \textit{High performance encapsulation in Casanova 2} , Computer Science and Electronic Engineering Conference (CEEC), 2015 7th, IEEE, pp. 201-206.
\item Abbadi, Mohamed; Di Giacomo, Francesco; Cortesi, Agostino; Pieter, Spronck; Giulia, Costantini; Giuseppe, Maggiore \textit{Casanova: A simple, high-performance language for game development} , Serious Games - First Joint International Conference, JCSG 2015, Springer, vol. 9090, pp. 123-134.
\item Di Giacomo, Francesco; Abbadi, Mohamed; Cortesi, Agostino; Spronck, Pieter; Maggiore, Giuseppe \textit{Building Game Scripting DSL’s with the Metacasanova Metacompiler}, Intelligent Technologies for Interactive Entertainment, Springer, vol. 178.
\item Di Giacomo, Francesco; Abbadi, Mohamed; Cortesi, Agostino; Spronck, Pieter; Maggiore, Giuseppe \textit{Metacasanova: an optimized meta-compiler for Domain-Specific Languages} , Proceedings of the 10th ACM SIGPLAN International Conference on Software Language Engineering.
\item Di Giacomo, Francesco; Abbadi, Mohamed; Cortesi, Agostino; Spronck, Pieter; Costantini, Giulia; Maggiore, Giuseppe \textit{High performance encapsulation and networking in Casanova 2} in Entertainment Computing, vol. 20, pp. 25-41.
\end{enumerate}