\begin{Ticclist}{}

\Ticcitem{Pashiera Barkhuysen.}{Audiovisual Prosody in Interaction.}{Promotores: M.G.J. Swerts, E.J. Krahmer.}{Tilburg, 3 October 2008.}

\Ticcitem{Ben Torben-Nielsen.}{Dendritic Morphology: Function Shapes Structure.}{Promotores: H.J. van den Herik, E.O. Postma. Co-promotor: K.P. Tuyls.}{Tilburg, 3 December 2008.}

\Ticcitem{Hans Stol.}{ A Framework for Evidence-based Policy Making Using IT.}{Promotor: H.J. van den Herik.}{ Tilburg, 21 January 2009.}

\Ticcitem{Jeroen Geertzen.}{ Dialogue Act Recognition and Prediction.}{ Promotor: H. Bunt. Co-promotor: J.M.B. Terken.}{Tilburg, 11 February 2009.}

\Ticcitem{Sander Canisius.}{ Structured Prediction for Natural Language Processing.}{Promotores: A.P.J. van den Bosch, W. Daelemans.}{ Tilburg, 13 February 2009.}

\Ticcitem{Fritz Reul.}{New Architectures in Computer Chess.}{ Promotor: H.J. van den Herik. Co-promotor: J.W.H.M. Uiterwijk.}{Tilburg, 17 June 2009.}

\Ticcitem{Laurens van der Maaten.}{ Feature Extraction from Visual Data.}{Promotores: E.O. Postma, H.J. van den Herik. Co-promotor: A.G. Lange.}{Tilburg, 23 June 2009 (cum laude).}

\Ticcitem{Stephan Raaijmakers.}{Multinomial Language Learning.}{ Promotores: W. Daelemans, A.P.J. van den Bosch.}{Tilburg, 1 December 2009.}

\Ticcitem{Igor Berezhnoy.}{ Digital Analysis of Paintings.}{ Promotores: E.O. Postma, H.J. van den Herik.}{Tilburg, 7 December 2009.}

\Ticcitem{Toine Bogers.}{ Recommender Systems for Social Bookmarking.}{ Promotor: A.P.J. van den Bosch.}{Tilburg, 8 December 2009.}

\Ticcitem{Sander Bakkes.}{ Rapid Adaptation of Video Game AI.}{ Promotor: H.J. van den Herik. Co-promotor: P. Spronck.}{Tilburg, 3 March 2010.}

\Ticcitem{Maria Mos.}{Complex Lexical Items.}{ Promotor: A.P.J. van den Bosch. Co-promotores: A. Vermeer, A. Backus.}{Tilburg, 12 May 2010 (in collaboration with the Department of Language and Culture Studies).}

\Ticcitem{Marieke van Erp.}{Accessing Natural History. Discoveries in data cleaning, structuring, and retrieval.}{Promotor: A.P.J. van den Bosch. Co-promotor: P.K. Lendvai.}{ Tilburg, 30 June 2010.}

\Ticcitem{Edwin Commandeur.}{Implicit Causality and Implicit Consequentiality in Language Comprehension.}{ Promotores: L.G.M. Noordman, W. Vonk. Co-promotor: R. Cozijn.}{ Tilburg, 30 June 2010.}

\Ticcitem{Bart Bogaert.}{Cloud Content Contention.}{ Promotores: H.J. van den Herik, E.O. Postma.}{ Tilburg, 30 March 2011.}

\Ticcitem{Xiaoyu Mao.}{Airport under Control.}{ Promotores: H.J. van den Herik, E.O. Postma. Co-promotores: N. Roos, A. Salden.}{ Tilburg, 25 May 2011.}

\Ticcitem{Olga Petukhova.}{Multidimensional Dialogue Modelling.}{Promotor: H. Bunt.}{Tilburg, 1 September 2011.}

\Ticcitem{Lisette Mol.}{Language in the Hands.}{Promotores: E.J. Krahmer, A.A. Maes, M.G.J. Swerts.}{Tilburg, 7 November 2011 (cum laude).}

\Ticcitem{Herman Stehouwer.}{Statistical Language Models for Alternative Sequence Selection.}{ Promotores: A.P.J. van den Bosch, H.J. van den Herik. Co-promotor: M.M. van Zaanen.}{Tilburg, 7 December 2011.}

\Ticcitem{Terry Kakeeto-Aelen.}{Relationship Marketing for SMEs in Uganda.}{Promotores: J. Chr. van Dalen, H.J. van den Herik. Co-promotor: B.A. Van de Walle.}{Tilburg, 1 February 2012.}

\Ticcitem{Suleman Shahid.}{Fun \& Face: Exploring non-verbal expressions of emotion during playful interactions.}{Promotores: E.J. Krahmer, M.G.J. Swerts.}{Tilburg, 25 May 2012.}

\Ticcitem{Thijs Vis.}{Intelligence, Politie en Veiligheidsdienst: Verenigbare Grootheden?}{ Promotores: T.A. de Roos, H.J. van den Herik, A.C.M. Spapens.}{Tilburg, 6 June 2012 (in collaboration with the Tilburg School of Law).}

\Ticcitem{Nancy Pascall.}{Engendering Technology Empowering Women.}{ Promotores: H.J. van den Herik, M. Diocaretz.}{Tilburg, 19 November 2012.}

\Ticcitem{Agus Gunawan.}{Information Access for SMEs in Indonesia.}{Promotor: H.J. van den Herik. Co-promotores: M. Wahdan, B.A. Van de Walle.}{Tilburg, 19 December 2012.}

\Ticcitem{Giel van Lankveld.}{Quantifying Individual Player Differences.}{Promotores: H.J. van den Herik, A.R. Arntz. Co-promotor: P. Spronck.}{ Tilburg, 27 February 2013.}

\Ticcitem{Sander Wubben.}{Text-to-text Generation Using Monolingual Machine Translation.}{ Promotores: E.J. Krahmer, A.P.J. van den Bosch, H. Bunt.}{Tilburg, 5 June 2013.}

\Ticcitem{Jeroen Janssens.}{Outlier Selection and One-Class Classification.}{Promotores: E.O. Postma, H.J. van den Herik.}{Tilburg, 11 June 2013.}

\Ticcitem{Martijn Balsters.}{Expression and Perception of Emotions: The Case of Depression, Sadness and Fear.}{Promotores: E.J. Krahmer, M.G.J. Swerts, A.J.J.M. Vingerhoets.}{Tilburg, 25 June 2013.}

\Ticcitem{Lisanne van Weelden.}{Metaphor in Good Shape.}{ Promotor: A.A. Maes. Co-promotor: J. Schilperoord.}{Tilburg, 28 June 2013.}

\Ticcitem{Ruud Koolen.}{“Need I say More? On Overspecification in Definite Reference.”}{Promotores: E.J. Krahmer and M.G.J. Swerts.}{Tilburg, 20 September 2013.}

\Ticcitem{J. Douglas Mastin.}{Exploring Infant Engagement. Language Socialization and Vocabulary. Development: A Study of Rural and Urban Communities in Mozambique.}{ Promotor: A.A. Maes. Co-promotor: P.A. Vogt.}{Tilburg, 11 October 2013.}

\Ticcitem{Philip C. Jackson. Jr.}{Toward Human-Level Artificial Intelligence – Representation and Computation of Meaning in Natural Language.}{Promotores: H.C. Bunt and W.P.M. Daelemans.}{Tilburg, 22 April 2014.}

\Ticcitem{Jorrig Vogels.}{Referential choices in language production: The Role of Accessibility.}{Promotores: A.A. Maes and E.J. Krahmer.}{ Tilburg, 23 April 2014.}

\Ticcitem{Peter de Kock.}{Anticipating Criminal Behaviour.}{Promotores: H.J. van den Herik and J.C. Scholtes. Co-promotor: P. Spronck.}{Tilburg, 10 September 2014.}

\Ticcitem{Constantijn Kaland.}{Prosodic marking of semantic contrasts: do speakers adapt to addressees?}{ Promotores: M.G.J. Swerts and E.J. Krahmer.}{Tilburg, 1 October 2014.}

\Ticcitem{Jasmina Marić.}{Web Communities, Immigration and Social Capital.}{ Promotor: H.J. van den Herik. Co-promotores: R. Cozijn and M. Spotti.}{Tilburg, 18 November 2014.}

\Ticcitem{Pauline Meesters.}{Intelligent Blauw.}{Promotores: H.J. van den Herik and T.A. de Roos.}{ Tilburg, 1 December 2014.}

\Ticcitem{Mandy Visser.}{Better use your head. How people learn to signal emotions in social contexts.}{Promotores: M.G.J. Swerts and E.J. Krahmer.}{Tilburg, 10 June 2015.}

\Ticcitem{Sterling Hutchinson.}{How symbolic and embodied representations work in concert.}{ Promotores: M.M. Louwerse and E.O. Postma.}{ Tilburg, 30 June 2015.}

\Ticcitem{Marieke Hoetjes.}{Talking hands. Reference in speech, gesture and sign.}{Promotores: E.J. Krahmer and M.G.J. Swerts.}{Tilburg, 7 October 2015}

\Ticcitem{Elisabeth Lubinga.}{Stop HIV. Start talking? The effects of rhetorical figures in health messages on conversations among South African adolescents.}{Promotores: A.A. Maes and C.J.M. Jansen.}{ Tilburg, 16 October 2015.}

\Ticcitem{Janet Bagorogoza.}{Knowledge Management and High Performance. The Uganda Financial Institutions Models for HPO.}{Promotores: H.J. van den Herik and B. van der Walle.}{ Tilburg, 24 November 2015.}

\Ticcitem{Hans Westerbeek.}{Visual realism: Exploring effects on memory, language production, comprehension, and preference.}{Promotores: A.A. Maes and M.G.J. Swerts. Co-promotor: M.A.A. van Amelsvoort.}{Tilburg, 10 February 2016.}

\Ticcitem{Matje van de Camp.}{A link to the Past: Constructing Historical Social Networks from Unstructured Data.}{Promotores: A.P.J. van den Bosch and E.O. Postma.}{ Tilburg, 2 Maart 2016.}

\Ticcitem{Annemarie Quispel.}{ Data for all: Data for all: How professionals and non-professionals in design use and evaluate information visualizations.}{Promotor: A.A. Maes. Co-promotor: J. Schilperoord.}{ Tilburg, 15 June 2016.}

\Ticcitem{Rick Tillman.}{Language Matters: The Influence of Language and Language Use on Cognition.}{Promotores: M.M. Louwerse and E.O. Postma.}{Tilburg, 30 June 2016.}

\Ticcitem{Ruud Mattheij.}{The Eyes Have It.}{Promoteres: E.O. Postma, H. J. Van den Herik and P.H.M. Spronck.}{Tilburg, 5 October 2016.}

\Ticcitem{Marten Pijl.}{Tracking of human motion over time.}{ Promotores: E. H. L. Aarts and M. M. Louwerse, Co-promotor: J. H. M. Korst.}{Tilburg, 14 December 2016.}

\Ticcitem{Yevgen Matusevych.}{Learning constructions from bilingual exposure: Computational studies of argument structure acquisition.}{ Promotor: A.M. Backus. Co-promotor: A.Alishahi.}{ Tilburg, 19 December 2016.}

\Ticcitem{Karin van Nispen.}{What can people with aphasia communicate with their hands? A study of representation techniques in pantomime and co-speech gesture.}{Promotor: E.J. Krahmer. Co-promotor: M. van de Sandt-Koenderman.}{Tilburg, 19 December 2016.}

\Ticcitem{Adriana Baltaretu.}{Speaking of landmarks. How visual information inuences reference in spatial domains.}{ Promotores: A.A. Maes and E.J. Krahmer.}{Tilburg, 22 December 2016.}

\Ticcitem{Mohamed Abbadi.}{Casanova 2, a domain specific language for general game development.}{Promotores: A.A. Maes, P.H.M. Spronck and A. Cortesi. Co-promotor: G. Maggiore.}{Tilburg, 10 March 2017.}

\Ticcitem{Shoshannah Tekofsky.}{You Are Who You Play You Are. Modelling Player Traits from Video Game Behavior.}{Promotores: E.O. Postma and P.H.M. Spronck.}{  Tilburg, 19 June 2017.}

\Ticcitem{Adel Alhuraibi.}{From IT-BusinessStrategic Alignment to Performance: A Moderated Mediation Model of Social Innovation, and Enterprise Governance of IT.}{Promotores: H.J. van den Herik and Prof. dr. B.A. Van de Walle. Co-promotor: Dr. S. Ankolekar.}{Tilburg, 26 September 2017.}

\Ticcitem{Wilma Latuny.}{The Power of Facial Expressions.}{Promotores: E.O. Postma and H.J. van den Herik.}{Tilburg, 29 September 2017.}

\Ticcitem{Sylvia Huwae.}{Different Cultures, Different Selves? Suppression of Emotions and Reactions to Transgressions across Cultures.}{Promotores: E.J. Krahmer and J. Schaafsma.}{Tilburg, 11 October, 2017.}

\Ticcitem{Mariana Serras Pereira.}{A Multimodal Approach to Children’s Deceptive Behavior.}{ Promotor: M. Swerts. Co-promotor: S. Shahid.}{Tilburg, 10 January, 2018.}

\Ticcitem{Emmelyn Croes.}{Meeting Face-to-Face Online: The Effects of Video-Mediated Communication on Relationship Formation.}{Promotores: E.J. Krahmer and M. Antheunis. Co-promotor A.P. Schouten.}{Tilburg, 28 March 2018.}

\Ticcitem{Lieke van Maastricht.}{Second Language Prosody: Intonation and Rhythm in Production and Perception.}{Promotores: E.J. Krahmer and M. Swerts.}{Tilburg, 9 May 2018.}

\Ticcitem{Nanne van Noord.}{Learning visual representations of style.}{Promotores: E.O. Postma and M. Louwerse.}{Tilburg, 16 May 2018.}

\Ticcitem{Ingrid Masson Carro.}{ Handmade: On the Cognitive Origins of Gestural Representations.}{ Promotores: E.J. Krahmer and M. Goudbeek.}{ Tilburg, 25 June 2018.}

\Ticcitem{Bart Joosten.}{Detecting Social Signals with Spatiotemporal Gabor Filters.}{Promotores: E.J. Krahmer and E.O. Postma.}{Tilburg, 29 June 2018.}

\Ticcitem{Yan Gu.}{Chinese hands of time: The effects of language and culture on temporal gestures and spatio-temporal reasoning.}{Promotor: M. Swerts. Co-promotores: M.W. Hoetjes, R. Cozijn.}{Tilburg, 5 June 2018.}

\Ticcitem{Thiago Castro Ferreira.}{Advances in Natural Language Generation: Generating Varied Outputs from Semantice Inputs.}{ Promotor: E.J. Krahmer. Co-promotor: S. Wubben.}{Tilburg, 19 September 2018.}

\Ticcitem{Yu Gu.}{Automatic Emotion Recognition from Mandarin Speech.}{Promotores: E.O. Postma, H.J. van den Herik and H.X. Lin.}{Tilburg, 28 November 2018.}

\Ticcitem{Francesco Di Giacomo.}{\thetitle}{Promotores: A. Cortesi, E.O. Postma, P.H.M. Spronck. Co-promotor: G. Maggiore.}{ Tilburg, 19 November 2018.}


\end{Ticclist}

